---
title: "Model"
page-layout: full
execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(parsnip)
library(broom)
library(gt)

# Load Louisiana wide dataset
data_path_la <- if (file.exists("data/clean/louisiana_wide_clean.csv")) {
  "data/clean/louisiana_wide_clean.csv"
} else {
  "../data/clean/louisiana_wide_clean.csv"
}

la_wide <- read_csv(data_path_la, show_col_types = FALSE) |>
  mutate(period = factor(period, levels = c("pre_ban", "post_ban")))

# Load South Carolina wide dataset
data_path_sc <- if (file.exists("data/clean/south_carolina_wide_clean.csv")) {
  "data/clean/south_carolina_wide_clean.csv"
} else {
  "../data/clean/south_carolina_wide_clean.csv"
}

sc_wide <- read_csv(data_path_sc, show_col_types = FALSE) |>
  mutate(period = factor(period, levels = c("pre_ban", "post_ban")))

# Load South Carolina district dataset (SC READY)
data_path_sc_dist <- if (file.exists("data/clean/sc_district_model_clean.csv")) {
  "data/clean/sc_district_model_clean.csv"
} else {
  "../data/clean/sc_district_model_clean.csv"
}

sc_district <- read_csv(data_path_sc_dist, show_col_types = FALSE) |>
  mutate(period = factor(period, levels = c("pre_ban", "post_ban")))

# Helper function to fit model and return gt table
fit_model_table <- function(col_name, title_text, data = la_wide) {
  formula_obj <- as.formula(paste(col_name, "~ period"))
  
  res <- linear_reg() |>
    set_engine("lm") |>
    fit(formula_obj, data = data) |>
    tidy(conf.int = TRUE) |>
    mutate(
      term = if_else(term == "(Intercept)", "Pre-Ban (Intercept)", "Post-Ban Change")
    ) |>
    select(
      Variable = term,
      Estimate = estimate,
      `95% CI Lower` = conf.low,
      `95% CI Upper` = conf.high
    )
  
  is_sig <- (res$`95% CI Lower`[2] > 0 || res$`95% CI Upper`[2] < 0)
  bg_color <- if (is_sig) "#e6f9e6" else "#f2f4f4"
  txt_color <- if (is_sig) "#145a32" else "#5d6d7e"
  font_wt <- if (is_sig) "bold" else "normal"
  
  res |>
    gt() |>
    fmt_number(columns = 2:4, decimals = 2) |>
    tab_style(
      style = list(
        cell_fill(color = bg_color),
        cell_text(color = txt_color, weight = font_wt)
      ),
      locations = cells_body(rows = 2)
    ) |>
    tab_options(table.width = pct(100)) |>
    tab_header(
      title = title_text
    )
}
```


## Proficiency Rate as a Function of Subject With Bans
### Model equations: grades 3–8 all subjects

$$
\text{Louisiana: } \widehat{\text{Proficiency Rate}} = 29.27 + 0.76 \cdot \text{PostBan}
$$

$$
\text{South Carolina: } \widehat{\text{Proficiency Rate}} = 39.44 + 3.63 \cdot \text{PostBan}
$$

The estimate of the intercept is the predicted score for that row. The other rows estimates show the net difference between that row and the intercept. So, a positive number means it is that much more than the intercept, and a negative means that much less.

```{r}
#| label: fig-model-predictions
#| fig-cap: "Model-predicted proficiency rates before and after cell phone bans for Louisiana (warm colors) and South Carolina (cold colors) across subjects."

library(readr)
library(dplyr)
library(parsnip)
library(broom)
library(marginaleffects)
library(ggplot2)

# Load cleaned datasets
data_path_la <- if (file.exists("data/clean/louisiana_wide_clean.csv")) {
  "data/clean/louisiana_wide_clean.csv"
} else {
  "../data/clean/louisiana_wide_clean.csv"
}

data_path_sc_dist <- if (file.exists("data/clean/sc_district_model_clean.csv")) {
  "data/clean/sc_district_model_clean.csv"
} else {
  "../data/clean/sc_district_model_clean.csv"
}

la_wide <- read_csv(data_path_la, show_col_types = FALSE) |>
  mutate(period = factor(period, levels = c("pre_ban", "post_ban")))

sc_district <- read_csv(data_path_sc_dist, show_col_types = FALSE) |>
  mutate(period = factor(period, levels = c("pre_ban", "post_ban")))

subjects_la <- list(
  list(col = "g3_8_all_subjects", name = "All Subjects"),
  list(col = "g3_8_ela", name = "ELA"),
  list(col = "g3_8_math", name = "Math"),
  list(col = "g3_8_science", name = "Science"),
  list(col = "g3_8_social_studies", name = "Social Studies")
)

subjects_sc <- list(
  list(col = "g3_8_all_subjects", name = "All Subjects"),
  list(col = "g3_8_ela", name = "ELA"),
  list(col = "g3_8_math", name = "Math")
)

preds_list <- list()

for (s in subjects_la) {
  m <- lm(as.formula(paste(s$col, "~ period")), data = la_wide)
  p <- predictions(m, newdata = datagrid(period = c("pre_ban", "post_ban"))) |>
    as_tibble() |>
    mutate(state = "Louisiana", subject = s$name)
  preds_list[[length(preds_list) + 1]] <- p
}

for (s in subjects_sc) {
  m <- lm(as.formula(paste(s$col, "~ period")), data = sc_district)
  p <- predictions(m, newdata = datagrid(period = c("pre_ban", "post_ban"))) |>
    as_tibble() |>
    mutate(state = "South Carolina", subject = s$name)
  preds_list[[length(preds_list) + 1]] <- p
}

all_preds <- bind_rows(preds_list) |>
  mutate(
    period_label = if_else(period == "pre_ban", "Pre-Ban", "Post-Ban"),
    period_label = factor(period_label, levels = c("Pre-Ban", "Post-Ban")),
    group_label = paste(state, "-", subject),
    subject = factor(subject, levels = c("All Subjects", "ELA", "Math", "Science", "Social Studies"))
  )

color_palette <- c(
  "Louisiana - All Subjects"        = "#d35400",
  "Louisiana - ELA"                 = "#e67e22",
  "Louisiana - Math"                = "#c0392b",
  "Louisiana - Science"             = "#e74c3c",
  "Louisiana - Social Studies"      = "#d4ac0d",
  "South Carolina - All Subjects"   = "#1b4f72",
  "South Carolina - ELA"            = "#2980b9",
  "South Carolina - Math"           = "#16a085"
)

dodge <- position_dodge(width = 0.35)

ggplot(all_preds, aes(x = period_label, y = estimate, color = group_label, group = group_label, shape = subject)) +
  geom_line(position = dodge, linewidth = 1.2, alpha = 0.85) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = dodge, size = 0.9, linewidth = 1.1) +
  facet_wrap(~ state) +
  scale_color_manual(values = color_palette) +
  scale_shape_manual(values = c("All Subjects" = 16, "ELA" = 17, "Math" = 15, "Science" = 18, "Social Studies" = 8)) +
  labs(
    title = "Predicted Student Proficiency Rates (Grades 3-8)",
    subtitle = "Comparing Louisiana (Warm Palette) and South Carolina (Cold Palette) Pre- vs. Post-Ban",
    x = "Policy Period",
    y = "Predicted Proficiency Rate (%)",
    color = "State & Subject",
    shape = "Subject"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 17),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

## Grades 3–8 Models: State Comparison

Red = negative change

Grey = no statistical change

Green = positive change

### All subjects

::: {layout-ncol=2}
```{r}
#| label: fit-g3-8-all-la
fit_model_table("g3_8_all_subjects", "Louisiana: All Subjects (G3-8)")
```

```{r}
#| label: fit-sc-dist-g3-8-all
fit_model_table("g3_8_all_subjects", "South Carolina: All Subjects (G3-8)", data = sc_district)
```
:::

### ELA

::: {layout-ncol=2}
```{r}
#| label: fit-g3-8-ela-la
fit_model_table("g3_8_ela", "Louisiana: ELA (G3-8)")
```

```{r}
#| label: fit-sc-dist-g3-8-ela
fit_model_table("g3_8_ela", "South Carolina: ELA (G3-8)", data = sc_district)
```
:::

### Math

::: {layout-ncol=2}
```{r}
#| label: fit-g3-8-math-la
fit_model_table("g3_8_math", "Louisiana: Math (G3-8)")
```

```{r}
#| label: fit-sc-dist-g3-8-math
fit_model_table("g3_8_math", "South Carolina: Math (G3-8)", data = sc_district)
```
:::

### Science

::: {layout-ncol=2}
```{r}
#| label: fit-g3-8-science-la
fit_model_table("g3_8_science", "Louisiana: Science (G3-8)")
```

*No matching South Carolina district data available for Science.*
:::

### Social Studies

::: {layout-ncol=2}
```{r}
#| label: fit-g3-8-soc-la
fit_model_table("g3_8_social_studies", "Louisiana: Social Studies (G3-8)")
```

*No matching South Carolina district data available for Social Studies.*
:::
This is the only statistically significant difference as the confidence interval does not incude 0 change. The other subjects have (within a 95% certainty range) a non-zero chance that the impact is 0.

## Grades 3–12 Models (Louisiana)

```{r}
#| label: fit-g3-12-all
fit_model_table("g3_12_all_subjects", "Grades 3-12: All Subjects")
```

```{r}
#| label: fit-g3-12-ela
fit_model_table("g3_12_ela", "Grades 3-12: ELA")
```

```{r}
#| label: fit-g3-12-math
fit_model_table("g3_12_math", "Grades 3-12: Math")
```

```{r}
#| label: fit-g3-12-science
fit_model_table("g3_12_science", "Grades 3-12: Science")
```

```{r}
#| label: fit-g3-12-soc
fit_model_table("g3_12_social_studies", "Grades 3-12: Social Studies")
```
This is also the only viable model becuase you can say with 95% certainty that there is a positive effect.


## Cell Phone Ban Effects by Grade Group (Louisiana)

```{r}
#| label: fig-model-boxplot-scatter
#| fig-cap: "Combined boxplot and scatterplot distributions comparing the policy effect of cell phone bans across grade groups in Louisiana (warm palette)."
#| fig-width: 7
#| fig-height: 4.5
#| fig-align: "center"
#| out-width: "80%"

library(ggplot2)

# Calculate Louisiana parish-level ban effects for grade groups
la_parish_pre <- la_wide |>
  filter(g3_12system_code != "LA", period == "pre_ban") |>
  group_by(g3_12system_code, system_name) |>
  summarize(
    pre_g3_8 = mean(g3_8_all_subjects, na.rm = TRUE),
    pre_g3_12 = mean(g3_12_all_subjects, na.rm = TRUE),
    .groups = "drop"
  )

la_parish_post <- la_wide |>
  filter(g3_12system_code != "LA", period == "post_ban") |>
  group_by(g3_12system_code, system_name) |>
  summarize(
    post_g3_8 = mean(g3_8_all_subjects, na.rm = TRUE),
    post_g3_12 = mean(g3_12_all_subjects, na.rm = TRUE),
    .groups = "drop"
  )

la_effects_df <- inner_join(la_parish_pre, la_parish_post, by = c("g3_12system_code", "system_name")) |>
  mutate(
    `Grades 3-8` = post_g3_8 - pre_g3_8,
    `Grades 3-12` = post_g3_12 - pre_g3_12
  ) |>
  pivot_longer(
    cols = c(`Grades 3-8`, `Grades 3-12`),
    names_to = "grade_group",
    values_to = "ban_effect"
  ) |>
  mutate(
    grade_group = factor(grade_group, levels = c("Grades 3-8", "Grades 3-12"))
  )

mean_effects <- la_effects_df |>
  group_by(grade_group) |>
  summarize(
    avg_effect = mean(ban_effect, na.rm = TRUE),
    .groups = "drop"
  )

diff_val <- mean_effects$avg_effect[2] - mean_effects$avg_effect[1]

ggplot(la_effects_df, aes(x = grade_group, y = ban_effect, fill = grade_group, color = grade_group)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.9) +
  geom_boxplot(alpha = 0.45, outlier.shape = NA, width = 0.45) +
  geom_jitter(width = 0.18, size = 0.8, alpha = 0.35, shape = 16) +
  geom_point(data = mean_effects, aes(x = grade_group, y = avg_effect), color = "black", size = 2.5, shape = 18, inherit.aes = FALSE) +
  geom_text(data = mean_effects, aes(x = grade_group, y = avg_effect + 1.2, label = paste0("Mean: +", sprintf("%.2f", avg_effect), "%")), color = "#7e2a0c", fontface = "bold", size = 3.8, inherit.aes = FALSE) +
  annotate("segment", x = 1, xend = 2, y = 8.5, yend = 8.5, color = "#9e3606", linewidth = 0.7) +
  annotate("segment", x = 1, xend = 1, y = 8.2, yend = 8.5, color = "#9e3606", linewidth = 0.7) +
  annotate("segment", x = 2, xend = 2, y = 8.2, yend = 8.5, color = "#9e3606", linewidth = 0.7) +
  annotate("text", x = 1.5, y = 9.2, label = paste0("Difference = ", sprintf("%.2f", diff_val), "% pts"), color = "#9e3606", fontface = "bold", size = 3.9) +
  scale_fill_manual(values = c("Grades 3-8" = "#e67e22", "Grades 3-12" = "#d35400")) +
  scale_color_manual(values = c("Grades 3-8" = "#9e3606", "Grades 3-12" = "#7e2a0c")) +
  scale_y_continuous(limits = c(-8, 10.5), breaks = seq(-8, 10, by = 2)) +
  labs(
    title = "Louisiana: Cell Phone Ban Policy Impact Across Grade Groups",
    subtitle = "Parish-Level Post-Ban Change in Proficiency Rate (% Points)\nDashed line at 0 = No Policy Effect",
    x = "Grade Group / Age Group",
    y = "Ban Effect (% Point Change)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10.5, color = "gray20"),
    plot.title.position = "plot",
    legend.position = "none",
    panel.grid.minor = element_blank()
  )
```


```{r}
#| label: calc-ban-effects-by-grade

# Louisiana: Cell Phone Ban Effects by Grade Level within Subject
la_ban_specs <- tibble::tribble(
  ~Subject, ~`Grade Level`, ~col,
  "All Subjects", "Grades 3-8", "g3_8_all_subjects",
  "All Subjects", "Grades 3-12", "g3_12_all_subjects",
  "ELA", "Grades 3-8", "g3_8_ela",
  "ELA", "Grades 3-12", "g3_12_ela",
  "Math", "Grades 3-8", "g3_8_math",
  "Math", "Grades 3-12", "g3_12_math",
  "Science", "Grades 3-8", "g3_8_science",
  "Science", "Grades 3-12", "g3_12_science",
  "Social Studies", "Grades 3-8", "g3_8_social_studies",
  "Social Studies", "Grades 3-12", "g3_12_social_studies"
)

la_ban_rows <- list()
for (i in 1:nrow(la_ban_specs)) {
  s <- la_ban_specs$Subject[i]
  g <- la_ban_specs$`Grade Level`[i]
  c <- la_ban_specs$col[i]
  
  f <- as.formula(paste(c, "~ period"))
  m <- lm(f, data = la_wide)
  res <- tidy(m, conf.int = TRUE)
  
  b0 <- res |> filter(term == "(Intercept)")
  b1 <- res |> filter(term == "periodpost_ban")
  
  la_ban_rows[[i]] <- tibble(
    Subject = s,
    `Grade Level` = g,
    `Pre-Ban Intercept (%)` = b0$estimate,
    `Ban Effect (%)` = b1$estimate,
    `Std Error` = b1$std.error,
    `95% CI Lower` = b1$conf.low,
    `95% CI Upper` = b1$conf.high
  )
}
la_ban_effects_df <- bind_rows(la_ban_rows)

# South Carolina: Cell Phone Ban Effects by Grade Level within Subject
sc_ban_specs <- tibble::tribble(
  ~Subject, ~`Grade Level`, ~col,
  "All Subjects", "Grades 3-8", "g3_8_all_subjects",
  "All Subjects", "Elementary (G3-5)", "elem_all_subjects",
  "All Subjects", "Middle School (G6-8)", "middle_all_subjects",
  "English", "High School (English II)", "hs_english",
  "Math", "High School (Algebra I)", "hs_math"
)

sc_ban_rows <- list()
for (i in 1:nrow(sc_ban_specs)) {
  s <- sc_ban_specs$Subject[i]
  g <- sc_ban_specs$`Grade Level`[i]
  c <- sc_ban_specs$col[i]
  
  f <- as.formula(paste(c, "~ period"))
  m <- lm(f, data = sc_wide)
  res <- tidy(m, conf.int = TRUE)
  
  b0 <- res |> filter(term == "(Intercept)")
  b1 <- res |> filter(term == "periodpost_ban")
  
  sc_ban_rows[[i]] <- tibble(
    Subject = s,
    `Grade Level` = g,
    `Pre-Ban Intercept (%)` = b0$estimate,
    `Ban Effect (%)` = b1$estimate,
    `Std Error` = b1$std.error,
    `95% CI Lower` = b1$conf.low,
    `95% CI Upper` = b1$conf.high
  )
}
sc_ban_effects_df <- bind_rows(sc_ban_rows)
```
<details>
<summary style="font-size: 1.4em; font-weight: bold; cursor: pointer; color: #1b4f72; margin-top: 1.5em; margin-bottom: 0.5em;">
 Cell Phone Ban Effects by Grade Level & Subject (Click to Expand)
</summary>

<br>

::: {layout-ncol=2}
```{r}
#| label: tbl-la-ban-effects
la_sig_rows <- which(la_ban_effects_df$`95% CI Lower` > 0 | la_ban_effects_df$`95% CI Upper` < 0)
la_grey_rows <- setdiff(1:nrow(la_ban_effects_df), la_sig_rows)

la_tbl <- la_ban_effects_df |>
  gt() |>
  fmt_number(columns = 3:7, decimals = 2)

if (length(la_sig_rows) > 0) {
  la_tbl <- la_tbl |>
    tab_style(
      style = list(
        cell_fill(color = "#e6f9e6"),
        cell_text(color = "#145a32", weight = "bold")
      ),
      locations = cells_body(rows = la_sig_rows)
    )
}

if (length(la_grey_rows) > 0) {
  la_tbl <- la_tbl |>
    tab_style(
      style = list(
        cell_fill(color = "#f2f4f4"),
        cell_text(color = "#5d6d7e")
      ),
      locations = cells_body(rows = la_grey_rows)
    )
}

la_tbl |>
  tab_options(table.width = pct(100)) |>
  tab_header(
    title = "Louisiana: Ban Effect by Grade Level & Subject",
    subtitle = "Green = 95% CI Excludes 0 (Statistically Significant) | Grey = 95% CI Includes 0"
  )
```

```{r}
#| label: tbl-sc-ban-effects
sc_sig_rows <- which(sc_ban_effects_df$`95% CI Lower` > 0 | sc_ban_effects_df$`95% CI Upper` < 0)
sc_grey_rows <- setdiff(1:nrow(sc_ban_effects_df), sc_sig_rows)

sc_tbl <- sc_ban_effects_df |>
  gt() |>
  fmt_number(columns = 3:7, decimals = 2)

if (length(sc_sig_rows) > 0) {
  sc_tbl <- sc_tbl |>
    tab_style(
      style = list(
        cell_fill(color = "#e6f9e6"),
        cell_text(color = "#145a32", weight = "bold")
      ),
      locations = cells_body(rows = sc_sig_rows)
    )
}

if (length(sc_grey_rows) > 0) {
  sc_tbl <- sc_tbl |>
    tab_style(
      style = list(
        cell_fill(color = "#f2f4f4"),
        cell_text(color = "#5d6d7e")
      ),
      locations = cells_body(rows = sc_grey_rows)
    )
}

sc_tbl |>
  tab_options(table.width = pct(100)) |>
  tab_header(
    title = "South Carolina: Ban Effect by Grade Level & Subject",
    subtitle = "Green = 95% CI Excludes 0 (Statistically Significant) | Grey = 95% CI Includes 0"
  )
```
:::

</details>

